Understanding meta-analysis through data simulation with applications to power analysis

### 2023-07-03 documents/paper/paper.Rmd
fixing s and equal_effects naming

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity [@Borenstein2009-mo, pp.187-203]. More recently, location-scale models have been developed to inlcude predictors also on the residual heterogeneity [@Viechtbauer2022-rz]. Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis [@Borenstein2009-mo; @Schmid2022-xe; @Harrer2021-bz], to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process [@Ingalls2011-xt; @Gelman2020-tg, pp. 69-76; @Gelman2006-pc, pp. 155-176]. A recent paper by DeBruine and Barr [-@DeBruine2021-id], which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for fixed and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the fixed vs. random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language `r rcit`. We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository ([https://osf.io/54djn/](https://osf.io/54djn/)). Multivariate and multilevel models’ simulation examples and more details about the coding approach are available in the supplementary materials.

@@

```{r img-fixed-vs-random, echo=FALSE, fig.asp=1, fig.cap=get_caption()}
r_imgs$fixed_vs_random
```

@@

The *random-effects model* (RE) can be considered an extension of the *equal-effects model*. The *equal-effects model* assumes that the real effect is a single value. The random-effects model relaxes this assumption allowing the true effect size to vary across studies. For example, the difference between groups we are simulating could be influenced by the type of experiment or the participants' age.
Now $\theta$ is no longer a single value but a distribution of values. Due to the effect size being a distribution, we need to estimate both the mean and the variance. The parameter $\mu_{\theta}$ is mean of the distribution, interpreted as the average effect size across different true effect sizes and $\tau^{2}$ is the variance of the distribution interpreted as variability or heterogeneity of effect sizes. In practical terms, we now have two sources of variability: $\tau^{2}$, which express the real difference among effect sizes, and $\sigma_{i}^{2}$ which is the known sampling variance of each study as in the *equal-effects model*. We can easily extend Equation \@ref(eq:equal-effects-model1) with Equations \@ref(eq:random-effect-model1), \@ref(eq:random-effect-model2) and \@ref(eq:random-effect-model3).

\begin{align}
\begin{aligned}
y_i = \mu_{\theta} + \delta_i + \epsilon_i \\
(\#eq:random-effect-model1)
\end{aligned}
\end{align}

\begin{align}
\begin{aligned}
\delta_i \sim \mathcal{N}(0,\tau^2)
(\#eq:random-effect-model2)
\end{aligned}
\end{align}

\begin{align}
\begin{aligned}
\epsilon_i \sim \mathcal{N}(0,\sigma_{\epsilon_i}^{2})
(\#eq:random-effect-model3)
\end{aligned}
\end{align}

@@

```{r sim-random-effect}
set.seed(seed)
k <- 30 # number of studies
n <- 20 # number of participants per group, per study
mu <- 0.3 # real effect size
tau2 <- 0.2 # the effect size heterogeneity

@@

Then we can fit the *random-effects* meta-analysis model with the `rma` function and `method = "REML"`. Table \@ref(tab:res-random-effect) depicts the model results^[The `method = "REML"` is not the only method to estimate a random-effects model. See the `rma` documentation (https://wviechtb.github.io/metafor/reference/rma.uni.html#specifying-the-model) for an overview of other estimation methods] of the metafor package. We are now estimating two parameters: $\mu_{\theta}$ and $\tau^{2}$. As for the *equal-effects model*, increasing the number of studies and/or the number of participants in each study will improve estimation reducing the standard error.

```{r res-random-effect, echo = FALSE}
res_tidy <- rma_tidy(res)
rma_table(res_tidy$res, res_tidy$notes, caption = get_caption())
```

An important aspect of the *random-effects model* is the interplay between the heterogeneity ($\tau^{2}$) and the sampling variability ($\sigma_{i}^{2}$). As the number of studies $k$ increases, the estimation of $\mu_{\theta}$ and $\tau^{2}$ becomes more precise. Additionally, as each study\'s sample size decreases, each $\theta_{i}$ will be estimated with higher precision but as long as $\tau^{2} \neq 0$ there will be variability among effect sizes. In other words, increasing the sample size of each study or the number of studies will not affect the value of $\tau^{2}$ but only the estimation precision [see @Borenstein2009-mo, Chapter 16, Figure 16.6 for a clear explanation]. This can be easily demonstrated using the previous simulation, increasing each study\'s sample size. Figure \@ref(fig:random-effect-high-low-precision) depicts the same meta-analysis but with different precision in estimating the study $y_{i}$. Compared to Figure \@ref(fig:equal-effects-high-low-precision), increasing the sample size of primary studies improves the estimation of each study without reducing the between-studies heterogeneity.

```{r random-effect-high-low-precision, echo = FALSE, fig.asp=0.8, fig.cap=get_caption()}
set.seed(seed)
k <- 10 # number of studies
mu <- 0.3 # real effect size
tau2 <- 0.2
n <- c(30, 500)

@@

From Equation \@ref(eq:isquared) and Figure \@ref(fig:random-effect-high-low-precision) is clear that if each included study has a considerable sample size the total variability will be reduced, mainly driven by real heterogeneity ($\tau^2$). This is the crucial difference between the equal-effects and the random-effects model [see also @Borenstein2009-mo, pp. 117–122].

Given the interpretation of $I^{2}$ it is possible to simulate a meta-analysis fixing a certain $I^{2}$ value. The only caveat is fixing the $\tilde{v}$. When the sample size of each study is the same, $\tilde{v} = \sigma^2_i$. In the other case, $\tilde{v}$ needs to be calculated from sampling variances reported in Higgins and Thompson [-@Higgins2002-fh] and cannot be easily fixed a priori. However, with the assumption of homogeneous sample size across studies^[The homogeneous sample size of primary studies assumption is commonly used to calculate the statistical power [see @Borenstein2009-mo, Chapter 29]], we can solve Equation \@ref(eq:isquared) for $\tau^{2}$ obtaining the heterogeneity value associated with a certain $I^{2}$ as reported in Equation \@ref(eq:tau2-from-isquared). Table \@ref(tab:res-random-effect-i2) depicts the results of the random-effects model fixing the $I^{2}$ value. Choosing a meaningful $\tau^2$ for the simulation can be sometimes difficult. Beyond fixing the $I^2$ value, one could use a plausible value from the literature. For example van Erp and colleagues [@Van_Erp2017-cf] estimate an empirical $\tau^2$ distribution across several published meta-analyses.

@@

```{r sim-random-effect-i2}
set.seed(seed)
k <- 30 # number of studies
mu <- 0.3 # real effect size
n <- 30 # sample size per group, per study
I2 <- 0.6 # desired I2 value

@@

```{r res-random-effect-i2, echo = FALSE}
res_tidy <- rma_tidy(res)
rma_table(res_tidy$res, res_tidy$notes, caption = get_caption())
```

@@

A common example of meta-regression is by including a categorical predictor, including information about study-level features. In our example, a group of studies uses an online memory task while others use a standard lab-based task. The Equation \@ref(eq:random-effect-model1) can be easily extended for a meta-regression model by including a variable encoding the type of task (online vs. lab-based) and the expected difference between the two levels of the moderator (i.e., the *lab vs. online effect*). In regression terms (see equation \@ref(eq:meta-regression)), we could use a dummy variable ($X_{1}$) that takes the value of 0 for the *lab-based task* (*L*) and a value of 1 for an *online task* (*O*)^[The *dummy coding* (also known as treatment coding) is the default in R and `metafor` but other coding schemes could be used [see @Schad2020-ht for an overview of contrast coding schemes]]. Now we fix $\beta_{1}$ to be the *lab vs. online effect* (i.e., the expected mean difference between the two groups of studies) and the product between $\beta_{1}X_{1}$ will consider the *lab vs. online effect*. Crucially, despite $\tau^{2}$ is still the heterogeneity between effect sizes, now we need to fix $\tau^{2}$ considering that we included a moderator. In other terms, $\delta_i \sim \mathcal{N}(0, \tau_{r}^{2})$ where $\tau^2_r$ is the residual heterogeneity after including the moderator. We can describe our model using equation \@ref(eq:meta-regression-equations) according to the value of $X_{1i}$ (L = lab-based experiments, O = online experiments).

@@

(ref:res-random-effect) Summary of the simulated random-effects model: Compared to the equal-effects model, there are more parameters. The $\beta$ is the average effect ($\theta_{r}$) with the standard error, 95% confidence interval, and the Wald z-test. The $\tau^{2}$ is the estimated heterogeneity and the $I^{2}$ (explained in Section \@ref(re-model)) represents the percentage of total variability due to between-study heterogeneity.

(ref:random-effect-high-low-precision) Forest plots of two simulated *random-effects model*s. On the left, the simulated model has $nt,nc = 30$ for each included study, while on the right, the sample size for each study is $nt,nc = 500$. Compared to Figure \@ref(fig:equal-effects-high-low-precision), data were generated under a *random-effects model*. The estimated average effect is similar between the two scenarios regarding average effect and precision. However, compared to the *equal-effects* simulation, increasing the sample size of primary studies only affects the precision without reducing the true heterogeneity (i.e., $\tau^{2} \neq 0$).

(ref:res-random-effect-i2) Summary of the *random-effect* model fixing the $I^2$ value. Parameters are the same as described in Table \@ref(tab:res-random-effect).

@@

(ref:res-meta-reg-dummy) Summary of the *random-effect* model with a categorical predictor (lab vs online experiments). The `intercept` is the average effect for lab-based experiments and `exponline` is the difference between lab-based and online experiments. The $R^2$ is the percentage of explained heterogeneity and $\tau^2_r$ is the estimated residual heterogeneity. Other parameters are the same as the standard *random-effect* model (see Table \@ref(tab:res-random-effect)).

(ref:res-meta-reg-numerical-r2) Summary of the *random-effects* model fixing the $R^{2}$ value. The `intercept` is the effect size for the average `age` (given that age is mean-centered). The `age0` parameter is the slope between `age` and the effect size interpreted as an increase in effect size for a unit increase in the average age. Other parameters are the same as described in Table \@ref(tab:res-random-effect) and \@ref(tab:res-meta-reg-dummy).

>>>>

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity [@Borenstein2009-mo, pp.187-203]. More recently, location-scale models have been developed to inlcude predictors also on the residual heterogeneity [@Viechtbauer2022-rz]. Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis [@Borenstein2009-mo; @Schmid2022-xe; @Harrer2021-bz], to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process [@Ingalls2011-xt; @Gelman2020-tg, pp. 69-76; @Gelman2006-pc, pp. 155-176]. A recent paper by DeBruine and Barr [-@DeBruine2021-id], which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for equal and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the fixed vs. random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language `r rcit`. We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository ([https://osf.io/54djn/](https://osf.io/54djn/)). Simulation examples for multivariate and multilevel models and more details about the coding approach are available in the supplementary materials.

@@

```{r img-equal-vs-random, echo=FALSE, fig.asp=1, fig.cap=get_caption()}
r_imgs$equal_vs_random
```

@@

The *random-effects model* (RE) can be considered an extension of the *equal-effects model*. The *equal-effects model* assumes that the real effect is a single value. The random-effects model relaxes this assumption allowing the true effect size to vary across studies. For example, the difference between groups we are simulating could be influenced by the type of experiment or the participants' age.
Now $\theta$ is no longer a single value but a distribution of values. Due to the effect size being a distribution, we need to estimate both the mean and the variance. The parameter $\mu_{\theta}$ is mean of the distribution, interpreted as the average effect size across different true effect sizes and $\tau^{2}$ is the variance of the distribution interpreted as variability or heterogeneity of effect sizes. In practical terms, we now have two sources of variability: $\tau^{2}$, which express the real difference among effect sizes, and $\sigma_{i}^{2}$ which is the known sampling variance of each study as in the *equal-effects model*. We can easily extend Equation \@ref(eq:equal-effects-model1) with Equations \@ref(eq:random-effects-model1), \@ref(eq:random-effects-model2) and \@ref(eq:random-effects-model3).

\begin{align}
\begin{aligned}
y_i = \mu_{\theta} + \delta_i + \epsilon_i \\
(\#eq:random-effects-model1)
\end{aligned}
\end{align}

\begin{align}
\begin{aligned}
\delta_i \sim \mathcal{N}(0,\tau^2)
(\#eq:random-effects-model2)
\end{aligned}
\end{align}

\begin{align}
\begin{aligned}
\epsilon_i \sim \mathcal{N}(0,\sigma_{\epsilon_i}^{2})
(\#eq:random-effects-model3)
\end{aligned}
\end{align}

@@

```{r sim-random-effects}
set.seed(seed)
k <- 30 # number of studies
n <- 20 # number of participants per group, per study
mu <- 0.3 # real effect size
tau2 <- 0.2 # the effect size heterogeneity

@@

Then we can fit the *random-effects* meta-analysis model with the `rma` function and `method = "REML"`. Table \@ref(tab:res-random-effects) depicts the model results^[The `method = "REML"` is not the only method to estimate a random-effects model. See the `rma` documentation (https://wviechtb.github.io/metafor/reference/rma.uni.html#specifying-the-model) for an overview of other estimation methods] of the metafor package. We are now estimating two parameters: $\mu_{\theta}$ and $\tau^{2}$. As for the *equal-effects model*, increasing the number of studies and/or the number of participants in each study will improve estimation reducing the standard error.

```{r res-random-effects, echo = FALSE}
res_tidy <- rma_tidy(res)
rma_table(res_tidy$res, res_tidy$notes, caption = get_caption())
```

An important aspect of the *random-effects model* is the interplay between the heterogeneity ($\tau^{2}$) and the sampling variability ($\sigma_{i}^{2}$). As the number of studies $k$ increases, the estimation of $\mu_{\theta}$ and $\tau^{2}$ becomes more precise. Additionally, as each study\'s sample size decreases, each $\theta_{i}$ will be estimated with higher precision but as long as $\tau^{2} \neq 0$ there will be variability among effect sizes. In other words, increasing the sample size of each study or the number of studies will not affect the value of $\tau^{2}$ but only the estimation precision [see @Borenstein2009-mo, Chapter 16, Figure 16.6 for a clear explanation]. This can be easily demonstrated using the previous simulation, increasing each study\'s sample size. Figure \@ref(fig:random-effects-high-low-precision) depicts the same meta-analysis but with different precision in estimating the study $y_{i}$. Compared to Figure \@ref(fig:equal-effects-high-low-precision), increasing the sample size of primary studies improves the estimation of each study without reducing the between-studies heterogeneity.

```{r random-effects-high-low-precision, echo = FALSE, fig.asp=0.8, fig.cap=get_caption()}
set.seed(seed)
k <- 10 # number of studies
mu <- 0.3 # real effect size
tau2 <- 0.2
n <- c(30, 500)

@@

From Equation \@ref(eq:isquared) and Figure \@ref(fig:random-effects-high-low-precision) is clear that if each included study has a considerable sample size the total variability will be reduced, mainly driven by real heterogeneity ($\tau^2$). This is the crucial difference between the equal-effects and the random-effects model [see also @Borenstein2009-mo, pp. 117–122].

Given the interpretation of $I^{2}$ it is possible to simulate a meta-analysis fixing a certain $I^{2}$ value. The only caveat is fixing the $\tilde{v}$. When the sample size of each study is the same, $\tilde{v} = \sigma^2_i$. In the other case, $\tilde{v}$ needs to be calculated from sampling variances reported in Higgins and Thompson [-@Higgins2002-fh] and cannot be easily fixed a priori. However, with the assumption of homogeneous sample size across studies^[The homogeneous sample size of primary studies assumption is commonly used to calculate the statistical power [see @Borenstein2009-mo, Chapter 29]], we can solve Equation \@ref(eq:isquared) for $\tau^{2}$ obtaining the heterogeneity value associated with a certain $I^{2}$ as reported in Equation \@ref(eq:tau2-from-isquared). Table \@ref(tab:res-random-effects-i2) depicts the results of the random-effects model fixing the $I^{2}$ value. Choosing a meaningful $\tau^2$ for the simulation can be sometimes difficult. Beyond fixing the $I^2$ value, one could use a plausible value from the literature. For example van Erp and colleagues [@Van_Erp2017-cf] estimate an empirical $\tau^2$ distribution across several published meta-analyses.

@@

```{r sim-random-effects-i2}
set.seed(seed)
k <- 30 # number of studies
mu <- 0.3 # real effect size
n <- 30 # sample size per group, per study
I2 <- 0.6 # desired I2 value

@@

```{r res-random-effects-i2, echo = FALSE}
res_tidy <- rma_tidy(res)
rma_table(res_tidy$res, res_tidy$notes, caption = get_caption())
```

@@

A common example of meta-regression is by including a categorical predictor, including information about study-level features. In our example, a group of studies uses an online memory task while others use a standard lab-based task. The Equation \@ref(eq:random-effects-model1) can be easily extended for a meta-regression model by including a variable encoding the type of task (online vs. lab-based) and the expected difference between the two levels of the moderator (i.e., the *lab vs. online effect*). In regression terms (see equation \@ref(eq:meta-regression)), we could use a dummy variable ($X_{1}$) that takes the value of 0 for the *lab-based task* (*L*) and a value of 1 for an *online task* (*O*)^[The *dummy coding* (also known as treatment coding) is the default in R and `metafor` but other coding schemes could be used [see @Schad2020-ht for an overview of contrast coding schemes]]. Now we fix $\beta_{1}$ to be the *lab vs. online effect* (i.e., the expected mean difference between the two groups of studies) and the product between $\beta_{1}X_{1}$ will consider the *lab vs. online effect*. Crucially, despite $\tau^{2}$ is still the heterogeneity between effect sizes, now we need to fix $\tau^{2}$ considering that we included a moderator. In other terms, $\delta_i \sim \mathcal{N}(0, \tau_{r}^{2})$ where $\tau^2_r$ is the residual heterogeneity after including the moderator. We can describe our model using equation \@ref(eq:meta-regression-equations) according to the value of $X_{1i}$ (L = lab-based experiments, O = online experiments).

@@

(ref:res-random-effects) Summary of the simulated random-effects model: Compared to the equal-effects model, there are more parameters. The $\beta$ is the average effect ($\theta_{r}$) with the standard error, 95% confidence interval, and the Wald z-test. The $\tau^{2}$ is the estimated heterogeneity and the $I^{2}$ (explained in Section \@ref(re-model)) represents the percentage of total variability due to between-study heterogeneity.

(ref:random-effects-high-low-precision) Forest plots of two simulated *random-effects model*s. On the left, the simulated model has $nt,nc = 30$ for each included study, while on the right, the sample size for each study is $nt,nc = 500$. Compared to Figure \@ref(fig:equal-effects-high-low-precision), data were generated under a *random-effects model*. The estimated average effect is similar between the two scenarios regarding average effect and precision. However, compared to the *equal-effects* simulation, increasing the sample size of primary studies only affects the precision without reducing the true heterogeneity (i.e., $\tau^{2} \neq 0$).

(ref:res-random-effects-i2) Summary of the *random-effect* model fixing the $I^2$ value. Parameters are the same as described in Table \@ref(tab:res-random-effects).

@@

(ref:res-meta-reg-dummy) Summary of the *random-effect* model with a categorical predictor (lab vs online experiments). The `intercept` is the average effect for lab-based experiments and `exponline` is the difference between lab-based and online experiments. The $R^2$ is the percentage of explained heterogeneity and $\tau^2_r$ is the estimated residual heterogeneity. Other parameters are the same as the standard *random-effect* model (see Table \@ref(tab:res-random-effects)).

(ref:res-meta-reg-numerical-r2) Summary of the *random-effects* model fixing the $R^{2}$ value. The `intercept` is the effect size for the average `age` (given that age is mean-centered). The `age0` parameter is the slope between `age` and the effect size interpreted as an increase in effect size for a unit increase in the average age. Other parameters are the same as described in Table \@ref(tab:res-random-effects) and \@ref(tab:res-meta-reg-dummy).
